Forecasting Influenza Hospitalizations Using a Bayesian Hierarchical Nonlinear Model with Discrepancy

### 2025-08-14 manuscript_res.tex
almost finished edits

The contribution of this manuscript is to introduce a two component framework 
for modeling HHS hospitalization forecasts where hospitalization data and 
years of ILI data are used to inform forecast models. The first modeling 
component is a model of ILI data and the second is a model of hospitalization 
data with ILI as a predictive covariate.
Herein we use ILI models similar to those in Osthus et al. 
\cite[]{osthus2019dynamic} and Ulloa \cite[]{ulloa2019} for ILI forecasting.
The model of Osthus et al. \cite[]{osthus2019dynamic} is a combined data 
assimilation and statistical regression model which involves a compartmental 
model in a probabilistic framework. Their model also includes an additional 
component for capturing a systematic discrepancy between the deterministic 
part of the model and the actual data, an idea which was first introduced by 
Kennedy and O'Hagan \cite[]{kennedy2001bayesian}. The model in Ulloa 
\cite[]{ulloa2019} is a Bayesian hierarchical regression model with an 
underlying function intended to capture the trajectory of the seasonal 
ILI data. Herein, we provide a framework under which discrepancy modeling may 
be used along with a general function modeling ILI data, and we show the 
effectiveness of including discrepancy modeling during certain periods of 
the flu season.

In line with the newer FluSight standard of forecasting hospitalizations, we 
model hospitalizations as a linear function of ILI. Thus forecasts produced 
herein target flu hospitalizations and are a mapping of ILI forecasts to 
hospitalizations. This allows for ILI data from many seasons to be exploited 
and for ILI forecasts to assist in forecasting hospitalizations, which has 
fewer seasons of data than ILI. Several modeling schemes and their forecasts 
for the 2023 flu hospitalization season are compared, and it is shown that the 
modeling decisions produce good forecast results for different states or times 
during the flu season.

@@

Model comparison was done by calculating a proper scoring rule for each 
forecast.
Proper scoring rules are the 
current standard for comparing performance between probabilistic forecasts 
and selecting the best forecasts according to the notion of maximizing 
sharpness subject to (auto-)calibration 
\cite[]{gneiting2007probabilistic, tsyplakov2013evaluation}. Proper scoring 
rules are commonly used in forecast comparison and have the property that a 
forecaster is incentivized to be honest in the reporting of their forecasts 
\cite[]{gneiting2007strictly, gneiting2014probabilistic}.
The proper scoring rules primarily used in the FluSight competition and the 
COVID-19 Forecast Hub and which we will use herein are the negatively oriented
(smaller is better)
weighted interval score (WIS)
\cite[]{mathis2024evaluation, bracher2021evaluating}.
The WIS is used for scoring quantile or interval 
forecasts 
\cite[]{gneiting2007strictly, gneiting2014probabilistic, bracher2021evaluating} 
and is defined in (\ref{eq:wis2}) where $Q$ is a forecast represented by all 
included quantiles, $B$ is the number of intervals,  $y^*$ is the observed 
value targeted by the forecast, $w_0 = 1/2$ and $w_b = \alpha_b / 2$ are 
weights for each interval, and $\alpha_b$ is the nominal level of the $b^{th}$ 
interval where $b \in \{1, ..., B\}$. 
$IS_{\alpha}$ is the interval score (IS), a proper scoring rule for a 
single interval, defined in (\ref{eq:is}). Here 
$\vmathbb{1}\{\cdot\}$ is the indicator function.

@@

The forecasts used in this section and in the subsequent section follow the 
FluSight convention of being quantiles corresponding to the 23 given 
probabilities $\\(0.010, 0.025, 0.050, 0.100, 0.150, …, 0.950, 0.975, 0.990)$
making each forecast a set of 11 predictive intervals and a predictive median.
Figure \ref{fig:wis_by_week_horizon} gives three plots showing results of the 
study. Plot a) shows WIS scores averaged over each left out season and each 
of the 500 simulation replicates. The ASGD forecasts on average clearly 
outperformed the other models considered for almost each week. The ASGD does
especially well at week 20 which contains forecasts of the holiday week 22, 
showing the particular importance of modeling discrepancy at and around that 
week. The SIRD outperforms the SIR model, again showing the importance of 
modeling discrepancy, but it is still outperformed by the ASG and even for a
few weeks by the BASE and ARIMA models. Plot b) shows boxplots of all forecast
scores by model and separated by horizon forecasts of 1 to 4 weeks ahead. Here
also the ASGD appears to perform the best with the other variations of the 
nonlinear models not always outperforming the BASE or ARIMA models. 
To assess calibration, plot c)
shows the empirical coverage of each model for 11 predictive intervals from 
10\% to 98\% nominal levels. Each model is was undercalibrated, but the ASGD 
model is the closest to being well calibrated, especially at the hightest 
nominal levels. Table \ref{tab:sim_wis_res} shows summaries of the mean WIS and
coverage for 50\% and 95\% predictive intervals.

@@

In this section we apply the forecast models to make forecasts for the 2023 
flu season weekly hospitalizations and compare the results with those of 
forecasts submitted to FluSight. The scoring of the forecasts is in the 
context of the FluSight competition where each competing forecast was 
submitted as a set of 23 quantiles corresponding to given probabilities.
Forecasts of 1, 2, 3, and 4-week ahead hospitalization counts were requested, 
and forecasts were made at the state and national level. The first week of 
forecasting took place during the week of October 7, 2023, and the final week 
was the week of May 4, 2024 making 30 total weeks of forecasts.
 The same format was used during the 2021 and 2022 seasons and for the 
 COVID-19 Forecast Hub \cite[]{mathis2024evaluation, bracher2021evaluating}. 
% Nearly teams participated and forecasts from over 40 models were submitted throughout the duration of the season. 
 Primary scores for evaluating each forecast were the WIS and the 
 log-weighted intveral score (LWIS). The LWIS is defined the same as the WIS
 but where the log of the observation as well as the forecast quatiles are taken
 prior to calculating the score. We primarily use the LWIS in this section,
 though forecasting results in terms of the WIS may also be found in the 
 supplementary materials ??.

@@

For the
remainder of this section, focus will be made on forecasts made by coupling the
ASGD model with normal hospitalization model where ILI is a quadratic predictor.
We denote this model as ASGD\_NORM2. 
This decision was based on the ASGD's performance in the simulation study in 
section \ref{sec:simulation2} as well as an analysis of several other forecast
models including ASG, SIRD, SIR ILI models coupled with other variations of the
hospitalization model where the ASGD\_NORM2 model showed the best performance.
Details
of that analysis are found in the supplementary materials ??.

The FluSight 
models to which the ASGD\_NORM2 model are compared are the 
20 non-ensemble models
submitted to FluSight during the 2023 season.
The submitted forecasts may be found at \cite{mathis2023flusight}.
Among the models
was the FluSight baseline. The baseline forecast had as a median the most 
recently observed 
hospitalization count, and the forecast uncertainty was based on 
differences between 
previous hospitalizations. The baseline forecasts was
similar to the baseline forecast used in
previous flu forecasting seasons and in the COVID-19 hub 
\cite[]{mathis2024evaluation, Cramer2022-hub-dataset}.

Figure \ref{fig:lwis_cover_sum} shows forecast results for 2023 flu 
hospitalization forecasts for all 21 models being compared but with results
for the ASGD\_NORM2 and FluSight baseline models highlighted. 
For each week of the 2023 season, the  
LWIS averaged over location and horizon 
for each model is shown in a). Boxplots of LWIS for 1, 2, 3, and 4-week ahead
horizons for the ASGD\_NORM2 and baseline models are shown in b). And 
overall empirical coverage of the 11 predictive intervals for all models is 
shown in c) with ASGD\_NORM2 and the baseline model again highlighted. 
These plots show that over the season the ASGD\_NORM2 model tended to 
outperform most others in terms of LWIS, and in general is superior to the 
baseline model. The coverage plot shows that the ASGD\_NORM2 model is well 
calibrated at most levels, with slight overcoverage at the lower levels
up to about 80\%, and has better coverage than all other models. Similar results
for the WIS are shown in the supplementary materials ??.

@@

During the FluSight competition, most participating teams would occasionally 
miss a forecast submission for some week, location, or horizon. This makes a 
simple average of scores an imperfect measure of comparison, as forecasts
of certain targets included for one model may be excluded by another model.
Thus in figure \ref{fig:state_and_date_lwis} 
we compare the ASGD\_NORM2 model with all 
other models directly by comparing LWIS scores averaged over only shared
forecasts. Each tile in the plots represents the ratio of average ASGD\_NORM2
LWIS to that of one FluSight model by location or by week.
We call this the relative LWIS or RLWIS. A value smaller than
1 indicates the ASGD\_NORM2 model outperforms the other model with 
increasing outperformance as the value goes to 0. A value 
higher than 1 indicateds the ASGD\_NORM2 is outperformed. A grey tile indicates
the competing model had no forecasts matching the specified target.
Similar plots showing WIS scores are found in the supplementary materials ??.

Our model outperforms most other models with only the 
UGA\_flucast-INFLAenza model appearing to outperform ours. 
Table \ref{tab:fin_analysis_stats} gives summary results for mean and median
WIS and LWIS scores, predictive coverage, and relative RWIS and RLWIS scores
for each model. These summary values show the ASGD\_NORM2 as performing very
well compared to all other models, and outperforming all models except
UGA\_flucast-INFLAenza by RWIS and RLWIS. 
The takeaway of this section, like those of section \ref{sec:simulation2}, show
the combination of well performing model of ILI, the 
hierarchical Bayesian ASGD, with a 
straightforward hospitalization model lead to good forecasts which 
outperformed almost all other non-ensemble forecasts during of the 2023
FluSight competition.

@@

 In this manuscript we introduce a statistical modeling framework which allows 
 for the incorporation of several ILI forecast modeling methods. Specifically, 
 we built upon Osthus et al. \cite[]{osthus2019dynamic} and introduced a 
 framework for modeling ILI which includes the use of an arbitrary function 
 for modeling the main trajectory of ILI along with modeling the discrepancy. 
 We model flu hospitalizations by incorporating the ILI forecast model into a 
 model forecasting hospitalizations where hospitalization predictions are a 
 linear or quadratic function of ILI.

The simulation study in section \ref{sec:simulation2} suggests the ASG 
function in ILI modeling may slightly outperform the SIR model according to 
the LogS and CRPS scoring rules, but in the analysis of the 2023 flu season 
forecasts, the SIR model was overall superior. The results from both the 
simulation study and the real data analysis suggest that the addition of a 
discrepancy component in ILI modeling may improve forecasts especially near 
the holiday week between Christmas and New Years day. It should not be assumed 
that these conclusions may be generalized for all locations in the US, weeks 
of a season, or for future flu seasons. 
% Rather than deciding to use or not to use certain of the modeling techniques here, a more promising option may be to combine forecasts from various models into an ensemble forecast.

Forecasting the seasonal influenza outbreak remains a challenging task for 
forecasters. The general modeling framework in this manuscript is successful 
under diverse modeling conditions for all locations in the US and may 
contribute to future forecasting efforts. All forecast models were presented 
as separate from one another, but in the various locations and times of the 
flu season, different models perform better than others. To build on the work 
done here, a natural step forward would be to combine all or a few selected 
forecasts into an ensemble forecast. Such an ensemble may work to cancel out 
certain model biases or highlight model strengths, leading to more robust 
forecasts.

>>>>

The contribution of this manuscript is to introduce a two component 
Bayesian model framework 
for modeling HHS hospitalization forecasts where hospitalization data and 
years of ILI data are used to inform forecast models. The first modeling 
component is a hierarchical model of ILI data. 
The second is a model of hospitalization 
data with ILI as a predictive covariate.
Herein we use Bayesian ILI models similar to those in Osthus et al. 
\cite[]{osthus2019dynamic} and Ulloa \cite[]{ulloa2019} for ILI forecasting.
The model of Osthus et al. \cite[]{osthus2019dynamic} is a combined data 
assimilation and statistical regression model which involves a compartmental 
model in a probabilistic framework. Their model also includes an additional 
component for capturing a systematic discrepancy between the deterministic 
part of the model and the actual data, an idea which was first introduced by 
Kennedy and O'Hagan \cite[]{kennedy2001bayesian}. The model in Ulloa 
\cite[]{ulloa2019} is a Bayesian hierarchical regression model with an 
underlying function intended to capture the trajectory of the seasonal 
ILI data. Herein, we provide a framework under which discrepancy modeling may 
be used along with a general function modeling ILI data. Modeling discrepancy
introduces some challenges and requires thoughtful selection of model 
constraints and prior distributions to be used effectively.

In line with the newer FluSight standard of forecasting hospitalizations, we 
model hospitalizations as a linear function of ILI. The Bayesian scheme 
produces a posterior predictive distribution of hospitalizations, which is 
shown to be a reasonably well calibrated probabilistic forecast.
The flu hospitalization forecasts are a mapping of ILI forecasts to 
hospitalizations, and this allows for ILI data from many seasons to be exploited 
and for ILI forecasts to assist in forecasting hospitalizations.
At this state of national flu forecasting, using ILI data can be especially
useful since there are many more seasons of available ILI data than 
hospitalization data. 
The forecasts produced in this manuscript are compared with out of the box 
forecast models in a simulation study, and in a real data analysis the 
forecasts are compared with 20 other forecasts from the 2023 FluSight forecast
competition. The proposed models outperform both the out of the box methods, and
they also outperform all but one of the FluSight models.

@@

Model comparison was done by calculating a proper scoring rule for each 
forecast.
Proper scoring rules are the 
current standard for comparing performance between probabilistic forecasts 
and selecting the best forecasts according to the notion of maximizing 
sharpness subject to (auto-)calibration 
\cite[]{gneiting2007probabilistic, tsyplakov2013evaluation}. Proper scoring 
rules are commonly used in forecast comparison, and they are deisgned such 
that a 
forecaster is incentivized to be honest in the reporting of their forecasts 
\cite[]{gneiting2007strictly, gneiting2014probabilistic}.
The proper scoring rule primarily used in the FluSight competition and the 
COVID-19 Forecast Hub, and which is used in this simulation study,
is the negatively oriented
(smaller is better)
weighted interval score (WIS)
\cite[]{mathis2024evaluation, bracher2021evaluating}.
The WIS is used for scoring quantile or interval 
forecasts, or forecasts made up of predicitive intervals of several nominal 
levels
\cite[]{gneiting2007strictly, gneiting2014probabilistic, bracher2021evaluating}. 
The WIS defined in (\ref{eq:wis2}) where $Q$ is a forecast represented by all 
included quantiles, $B$ is the number of intervals,  $y^*$ is the observed 
value targeted by the forecast, $w_0 = 1/2$ and $w_b = \alpha_b / 2$ are 
weights for each interval, and $\alpha_b$ is the nominal level of the $b^{th}$ 
interval where $b \in \{1, ..., B\}$. 
$IS_{\alpha}$ is the interval score (IS), a proper scoring rule for a 
single interval, defined in (\ref{eq:is}). Here 
$\vmathbb{1}\{\cdot\}$ is the indicator function.

@@

The forecasts used in this section and in the subsequent section follow the 
FluSight convention of being quantiles corresponding to the 23 given 
probabilities $\\(0.010, 0.025, 0.050, 0.100, 0.150, …, 0.950, 0.975, 0.990)$
making each forecast a set of 11 predictive intervals and a predictive median.
Figure \ref{fig:wis_by_week_horizon} gives three plots showing results of the 
study. Plot a) shows WIS scores averaged over each left out season and each 
of the 500 simulation replicates. The ASGD forecasts on average clearly 
outperformed the other models considered in almost every week. The ASGD does
especially well at week 20 which contains forecasts of the holiday week 22. 
This demonstrates the
particular importance of modeling discrepancy at and around that 
week. The SIRD outperforms the SIR model, again showing the importance of 
modeling discrepancy, but it is still outperformed by the ASG and even for a
few weeks by the BASE and ARIMA models. Plot b) shows boxplots of all forecast
scores by model and separated by horizon forecasts of 1 to 4 weeks ahead. Here
also the ASGD performs the best with the other variations of the 
nonlinear models not always outperforming the BASE or ARIMA models. 
To assess calibration, plot c)
shows the empirical coverage of each model for 11 predictive intervals from 
10\% to 98\% nominal levels. Each model is undercalibrated, but the ASGD 
model is the closest to being well calibrated, especially at the hightest 
nominal levels. Table \ref{tab:sim_wis_res} shows summaries of the mean WIS and
coverage for 50\% and 95\% predictive intervals.

@@

In this section we apply the forecast models to make forecasts for the 2023 
flu season weekly hospitalizations and compare the results with those of 
forecasts submitted to FluSight. The scoring of the forecasts is in the 
context of the FluSight competition where each competing forecast was 
submitted as a set of 23 quantiles corresponding to given probabilities.
Forecasts of 1, 2, 3, and 4-week ahead hospitalization counts were requested, 
and forecasts were made at the state and national level. The first week of 
forecasting took place during the week of October 7, 2023, and the final week 
was the week of May 4, 2024 making 30 total weeks of forecasts.
 The same format was used during the 2021 and 2022 seasons and for the 
 COVID-19 Forecast Hub \cite[]{mathis2024evaluation, bracher2021evaluating}. 
% Nearly teams participated and forecasts from over 40 models were submitted throughout the duration of the season. 
 Primary scores for evaluating each forecast were the WIS and the 
 log-weighted intveral score (LWIS). The LWIS is defined the same as the WIS
 but where the log of the observation as well as the 
 log forecast quantiles are taken
 prior to calculating the score. We primarily use the LWIS in this section,
 though we including results for the WIS in the 
 supplementary materials.

@@

For the
remainder of this section, focus will be made on forecasts made by coupling the
ASGD and SIRD models with the 
normal hospitalization model where ILI is a quadratic predictor.
These forecast schemes are denoted as ASGD\_NORM2 and SIRD\_NOMR2. 
In the supplementary materials, we include analysis of 2023 hospitalization
forecasts for several other forecast
models including ASG, SIRD, SIR ILI models coupled with other variations of the
hospitalization model where the ASGD\_NORM2 and SIRD\_NORM2 models
showed the best and second best results respectively.

The FluSight 
models to which are used for comparison are
20 non-ensemble models
submitted to FluSight during the 2023 season.
The submitted forecasts may be found at \cite{mathis2023flusight}.
Among the models
was the FluSight baseline. The baseline forecast had as a median the most 
recently observed 
hospitalization count, and the forecast uncertainty was based on 
differences between 
previous hospitalizations. The baseline forecasts was
similar to the baseline forecast used in
previous flu forecasting seasons and in the COVID-19 hub 
\cite[]{mathis2024evaluation, Cramer2022-hub-dataset}.

Figure \ref{fig:lwis_cover_sum} shows forecast results for 2023 flu 
hospitalization forecasts for all 22 models being compared but with results
for the ASGD\_NORM2, SIRD\_NORM2, and FluSight baseline models highlighted. 
For each week of the 2023 season, the  
LWIS averaged over location and horizon 
for each model is shown in a). Boxplots of LWIS for 1, 2, 3, and 4-week ahead
horizons for the ASGD\_NORM2, SIRD\_NORM2, and baseline models are shown in b). 
And 
overall empirical coverage of the 11 predictive intervals for all models is 
shown in c) with ASGD\_NORM2, SIRD\_NORM2
and the baseline model again highlighted. 
These plots show that over the season the ASGD\_NORM2 model tended to 
outperform most others in terms of LWIS, and in general is superior to the 
baseline model. The coverage plot shows that the ASGD\_NORM2 model is well 
calibrated at most levels, with slight overcoverage at the lower levels
up to about 80\%, and has better coverage than all other models. Similar results
for the WIS are shown in the supplementary materials.

@@

During the FluSight competition, most participating teams would occasionally 
miss a forecast submission for some week, location, or horizon. This makes a 
simple average of scores an imperfect measure for comparison, as forecasts
of certain targets included for one model may be excluded by another model.
Thus in figure \ref{fig:state_and_date_lwis} 
the ASGD\_NORM2 model is compared with all 
other models individually using LWIS scores averaged over only the 
forecasts shared by the two models.
Each tile represents the ratio of average ASGD\_NORM2
LWIS to that of one FluSight model by location or by week.
We call this the relative LWIS or RLWIS. A value smaller than
1 indicates the ASGD\_NORM2 model outperforms the competing model. A value 
higher than 1 indicateds the ASGD\_NORM2 is outperformed. A grey tile indicates
the competing model had no forecasts matching the specified target.
Similar plots showing WIS scores are found in the supplementary materials. 
The ASGD\_NORM2 outperforms most other competing models more often than not, but
there are some weeks or locations where forecasting was poorer. For example, 
the plot shows the ASGD\_NORM2 performed poorly in Montana relative to the other
models.

Table \ref{tab:fin_analysis_stats} gives summary results for mean and median
WIS and LWIS scores, predictive coverage, and relative RWIS and RLWIS scores
for each model. Along with the plots in this section, it 
show that over the course of the 
2023 season, 
the ASGD\_NORM2 outperformed every other model except the 
UGA\_flucast-INFLAenza. The SIRD\_NORM2 is also one of the top performing 
forecast models.

@@

 In this manuscript we introduce a statistical modeling framework which allows 
 for the incorporation of several nonlinear
 ILI forecast modeling methods. Specifically, 
 we built upon Osthus et al. \cite[]{osthus2019dynamic} and introduced a 
 framework for modeling ILI which includes the use of an arbitrary function 
 for modeling the main trajectory of ILI along with modeling the discrepancy and
 specifically incorporate the smooth ASG function and compare forecasts using 
 the ASG with those using SIR. 
 We model flu hospitalizations by incorporating the ILI forecast model into a 
 model forecasting hospitalizations where hospitalization predictions are a 
 linear or quadratic function of ILI. Utlizing the ILI data which has been
 available for many more years than flu hospitalization, greatly improves
 one's ability to forecast hospitalizations, and the borrowing information 
 from all previous seasons through hierarhcical modeling is able to further
 exploit the ILI data. Including model iscrepancy was an important piece of 
 successful forecasting, but it required making thoughtful model constraints
 and selection of prior distributions.

The simulation study in section \ref{sec:simulation2} suggests the ASGD 
function in ILI modeling outperforms the SIRD model and out of the box 
forecast methods according to 
the WIS scoring rule, and in the analysis of 2023 FluSight forecasts the 
ASGD\_NORM2 modeling scheme outperformed all but one of the competing
forecasts when looking at the whole season. Though the ASGD\_NORM2 did very 
well across all seasons, for some weeks and locations its perfomance was not
as good. It should also not be assumed that ASGD\_NORM2 forecasts would perform
as well in every season given that each flu season will come with its own
challenges which the modeling scheme may or may not address.
% Rather than deciding to use or not to use certain of the modeling techniques here, a more promising option may be to combine forecasts from various models into an ensemble forecast.

The models in this manuscript were such that the data in each
location were modeled as
independent from all other locations. It is reasonable to assume that a
disease outbreak in one state will share characterstics with the outbreak in 
neighboring states, and indeed the behavior of ILI and hospitalizations is more
closely related for neighboring states as shown in data plots in the
supplementary materials. The one 2023 FluSight forecast model which outperformed
the ASGD\_NORM2, the UGA\_flucast-INFLAenza, was listed as being a spatial
model. And in past seasons, some of the top performing forecast models also 
explicitly modeled spatial relationships \cite[]{osthus2021multiscale}.
Likewise, the forecasts to which ours was compared were all non-ensemble 
forecasts which were all outperformed by the ensemble forecasts in FluSight.
Futher extensions of the models herein could include explicit modeling of 
spatial relationships between states and inclusion in an ensemble of other
forecast models.
